---
title: "Simple recipes for three basic Observable plot types in Quarto"
subtitle: "Spiced with some R"
author: "Martin C. Arnold"
format: 
  html:
    css: styles.css 
    code-fold: false
---

Click [here]() to see the source of this HTML document.

Observable offers many different ways to graph your data, but the line chart, bar chart, and scatter plot are basic and should definitely be learned. Tremendous flexibility and customisation are possible for such charts, but in this article we will cover only the simple cases--inspired by the ones presented [here](https://observablehq.com/@observablehq/chart-starter-recipes)--and focus on integrating them into a simple [Quarto document](https://quarto.org/docs/guide/) such that data can be manipulated and provided using R.

## Preliminaries

The `.qmd` file for generating this document has the following YAML header.

```{YAML}
---
title: "A line graph of Apple stock close prices"
author: "Martin C. Arnold"
format: 
  html:
    css: styles.css 
    code-fold: false
---
```

`styles.css` contains the following CSS code for centering SVG elements in their enclosing divs.

```{CSS}
div > svg {
  margin-left: auto;
  margin-right: auto;
}
```

::: {.column-margin}
Observable plots are shown in SVG elements. Centering these is personal preference and thus optional 🙂.
:::

We will make use of the below R packages.

```{r, warning=F, message=F}
library(tidyverse)
library(cowplot)
library(reactable)
```

## Line chart

We may to attach external data for Observable plots in a chunk with header ````{OJS}` and use the `FileAttachement()` function. Below ee use the `.csv` method since the we're dealing with a `.csv` file. The code below assumes that `aapl.csv` is in the same folder as the `.qmd` file. The `.csv` is attached with the observable notebook linked above.

::: {.column-margin}
Note that there's no need to include any JavaScript libraries in the document or chunk header--Quarto takes care of it!
:::

```{ojs, echo=T, eval=T}
/* in OJS we would do it like this  */
data = FileAttachment("aapl.csv")
  .csv({typed: true});
```

We may now graph the data. Note that in line we set dimensions of the SVG element in line 2 using `width: 600, height: 400`.

```{ojs}
//| eval: true
//| fig-cap: "Apple Stock -- close prices"

Plot.plot({
  width: 600, height: 400,
  marks: [
    Plot.ruleY([40]),
    Plot.lineY(data, {x: "Date", y: "Close"})
  ]
})
```
::: {.column-margin}
Here's an R version of the plot:

```{r, message = F}
R_data <- read_csv(
  "aapl.csv"
)
ggplot(R_data, 
  aes(x = Date, y = Close)
  ) + 
  geom_line() +
  theme_cowplot()
```
:::


## Bar chart

Similarly we may generate bar charts. Note that the dataset `alphabet` does not have to be included externally, because it is already available with the OJS library.

```{ojs}
//| eval: true
//| fig-cap: "Relative frequencies of letters in `alphabet` dataset"

Plot.plot({
  width: 600, height: 400,
  y: { label: "freq" },
  marks: [
    Plot.barY(
      alphabet, 
      {x: "letter", y: "frequency", sort: {x: "y", reverse: true}}
    ),
    Plot.ruleY([0])
  ]
})
```

## Scatterplot

The dataset `cars` (I think it's the same data as in the popular R dataset) also comes with the OJS library . We create the scatterplot as follows:

```{ojs}
//| eval: true
//| fig-cap: "Fuel economy of cars in `cars` dataset"

Plot.plot({
  width: 600, height: 400,
  marks: [
    Plot.dot(cars, {x: "power (hp)", y: "economy (mpg)"})
  ]
})
```

# Manipulating data with and supplying it using R   

Let's play with the `iris` data in `R` and then make it available to Observable.js (OJS). We first allow users to glimpse at the data using [reactable](https://glin.github.io/reactable/).

```{r, eval=T, echo=T, message=F}
reactable::reactable(iris)
```

Next, we summarise variables by species (by computing group means for all variables in the datset).

```{r}
iris_means <- iris %>% 
  group_by(Species) %>% 
  summarise_all(.funs = mean)
```

R-data is made available to OJS using `ojs_define()`.

```{r}
ojs_define(
  iris = iris,
  iris_means = iris_means
)
```

We may check that these data are indeed available to Observable: they are objects that hold multiple arrays. The following statements are evaluated in `ojs` chunks.

```{ojs}
iris
iris_means
```

For the data to be "compatible" with Observable plot we need to `transpose()`. Below we plot `Sepal.Length` and `Sepal.Width` colored by `Species` and add marks for their group means. 

```{ojs}
/* define plot object */
iris_plot = Plot.plot({
  width: 600, height: 400,
  marks: [
    Plot.dot(
      transpose(iris), {
        x: "Sepal.Length", 
        y: "Sepal.Width",
        fill: "Species",
        r: 3
      }),
      
    Plot.dot(transpose(iris_means), {
       x: "Sepal.Length", 
       y: "Sepal.Width",
       fill: "Species",
       r: 6
    })
  ]
})

/* add legend for color */  
iris_plot.legend("color")
```
